---
title: "reproducible research on Health and Economic Impacts of US Weather"
author: "Stella"
date: '2022-11-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

In this project, we analyze data on storms from the US National Oceanic and Atmospheric Administration (NOAA). We estimate the damage done by each type of event (Wind, Tornado, Storm, Snow, etc). We are trying to determine which event is most harmful to the US population and to the economy. Through the reseach, we can conclude that the Tornado is the most harmful to the US population and Flood is the most harmful to the US economy. 

## Data Processing Part 1: loading libraries
```{r load}
library(ggplot2)  
library(plyr) 
library(dplyr)


#load data
if(!exists("storm.data")) {
  storm.data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"),header = TRUE)
}
```

## Data Processing Part 2: Extrracting and Sorting Data
```{r sort}
vars <- c( "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
mydata <- storm.data[, vars]

tail(mydata)

#sorting
# create a new variable EVENT to transform variable EVTYPE in groups
mydata$EVENT <- "OTHER"
# group by keyword in EVTYPE
mydata$EVENT[grep("HAIL", mydata$EVTYPE, ignore.case = TRUE)] <- "HAIL"
mydata$EVENT[grep("HEAT", mydata$EVTYPE, ignore.case = TRUE)] <- "HEAT"
mydata$EVENT[grep("FLOOD", mydata$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
mydata$EVENT[grep("WIND", mydata$EVTYPE, ignore.case = TRUE)] <- "WIND"
mydata$EVENT[grep("STORM", mydata$EVTYPE, ignore.case = TRUE)] <- "STORM"
mydata$EVENT[grep("SNOW", mydata$EVTYPE, ignore.case = TRUE)] <- "SNOW"
mydata$EVENT[grep("TORNADO", mydata$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
mydata$EVENT[grep("WINTER", mydata$EVTYPE, ignore.case = TRUE)] <- "WINTER"
mydata$EVENT[grep("RAIN", mydata$EVTYPE, ignore.case = TRUE)] <- "RAIN"
# listing the transformed event types 
sort(table(mydata$EVENT), decreasing = TRUE)


mydata$PROPDMGEXP <- as.character(mydata$PROPDMGEXP)
mydata$PROPDMGEXP[is.na(mydata$PROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$PROPDMGEXP[!grepl("K|M|B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$PROPDMGEXP[grep("K", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$PROPDMGEXP[grep("M", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$PROPDMGEXP[grep("B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$PROPDMGEXP <- as.numeric(as.character(mydata$PROPDMGEXP))
mydata$property.damage <- mydata$PROPDMG * 10^mydata$PROPDMGEXP

mydata$CROPDMGEXP <- as.character(mydata$CROPDMGEXP)
mydata$CROPDMGEXP[is.na(mydata$CROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$CROPDMGEXP[!grepl("K|M|B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$CROPDMGEXP[grep("K", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$CROPDMGEXP[grep("M", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$CROPDMGEXP[grep("B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$CROPDMGEXP <- as.numeric(as.character(mydata$CROPDMGEXP))
mydata$crop.damage <- mydata$CROPDMG * 10^mydata$CROPDMGEXP

```
## Data Processing Part 3: Finding Aggregate Data
```{r find}
agg.fatalites.and.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES + INJURIES,  na.rm = TRUE))
agg.fatalites.and.injuries$type <- "fatalities and injuries"

# aggregate FATALITIES by type of EVENT
agg.fatalities <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES, na.rm = TRUE))
agg.fatalities$type <- "fatalities"

# aggregate INJURIES by type of EVENT
agg.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(INJURIES, na.rm = TRUE))
agg.injuries$type <- "injuries"

# combine all
agg.health <- rbind(agg.fatalities, agg.injuries)

health.by.event <- join (agg.fatalities, agg.injuries, by="EVENT", type="inner")


agg.propdmg.and.cropdmg <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage + crop.damage,  na.rm = TRUE))
agg.propdmg.and.cropdmg$type <- "property and crop damage"

# aggregate PropDamage by type of EVENT
agg.prop <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage, na.rm = TRUE))
agg.prop$type <- "property"

# aggregate INJURIES by type of EVENT
agg.crop <- ddply(mydata, .(EVENT), summarize, Total = sum(crop.damage, na.rm = TRUE))
agg.crop$type <- "crop"

# combine all
agg.economic <- rbind(agg.prop, agg.crop)


economic.by.event <- join (agg.prop, agg.crop, by="EVENT", type="inner")
```
## Results to question 1

```{r answerOne}
#Formatting results
# transform EVENT to factor variable for health variables
agg.health$EVENT <- as.factor(agg.health$EVENT)

# plot FATALITIES and INJURIES by EVENT
health.plot <- ggplot(agg.health, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total number of health impact") +
  ggtitle("Weather event types impact on public health") +
  theme(plot.title = element_text(hjust = 0.5))
print(health.plot)
```
## Results to question 2

```{r answerTwo}
# transform EVENT to factor variable for economic variables
agg.economic$EVENT <- as.factor(agg.economic$EVENT)

# plot PROPERTY damage and CROP damage by EVENT
economic.plot <- ggplot(agg.economic, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total damage in dollars") +
  ggtitle("Weather event types impact on property and crop damage") +
  theme(plot.title = element_text(hjust = 0.5))
print(economic.plot)
```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
